---
output:
  bookdown::html_document2:
    fig_caption: yes
editor_options:
  chunk_output_type: console
---

```{r echo = FALSE, cache = FALSE}
source("utils.R", local = TRUE)
```

R Basics {#CHAPTER-R-BASICS}
========

This chapter covers the basics: installing and using packages and loading data.

If you want to get started quickly, most of the recipes in this book require the ggplot2 and gcookbook packages to be installed on your computer. To do this, run:

```{r, eval=FALSE}
install.packages(c("ggplot2", "gcookbook"))
```

Then, in each R session, before running the examples in this book, you can load them with:

```{r, eval=FALSE}
library(ggplot2)
library(gcookbook)
```

> **Note**
>
> Chapter \@ref(CHAPTER-GGPLOT2) provides an introduction to the ggplot2 graphing package, for readers who are not already familiar with its use.

Packages in R are collections of functions and/or data that are bundled up for easy distribution, and installing a package will extend the functionality of R on your computer. If an R user creates a package and thinks that it might be useful for others, that user can distribute it through a package repository. The primary repository for distributing R packages is called CRAN (the Comprehensive R Archive Network), but there are others, such as Bioconductor and Omegahat.

Installing a Package
--------------------

### Problem

You want to install a package from CRAN.

### Solution

Use install.packages() and give it the name of the package you want to install. To install ggplot2, run:

```{r, eval=FALSE}
install.packages("ggplot2")
```

At this point you may be prompted to select a download mirror. It's usually best to use the first choice, https://cloud.r-project.org/, as it is a cloud-based mirror with endpoints all over the world.

### Discussion

When you tell R to install a package, it will automatically install any other packages that the first package depends on.

CRAN (the Comprehensive R Archive Network) is a repository of packages for R, and it is mirrored on many servers around the world. It is the default repository system used by R. There are other package repositories; Bioconductor, for example, is a repository of packages related to analyzing genomic data.

Loading a Package
-----------------

### Problem

You want to load an installed package.

### Solution

Use `library()` and give it the name of the package you want to install. To load ggplot2, run:

```{r, eval=FALSE}
library(ggplot2)
```

The package must already be installed on the computer.

### Discussion

Most of the recipes in this book require loading a package before running the code, either for the graphing capabilities (as in the ggplot2 package) or for example data sets (as in the MASS and gcookbook packages).

One of R's quirks is the package/library terminology. Although you use the `library()` function to load a package, a package is not a library, and some longtime R users will get irate if you call it that.

A *library* is a directory that contains a set of packages. You might, for example, have a system-wide library as well as a library for each user.

Loading a Delimited Text Data File
----------------------------------

### Problem

You want to load data from a delimited text file.

### Solution

The most common way to read in a file is to use comma-separated values (CSV) data:

```{r, eval=FALSE}
data <- read.csv("datafile.csv")
```

### Discussion

Since data files have many different formats, there are many options for loading them. For example, if the data file does *not* have headers in the first row:

```{r, eval=FALSE}
data <- read.csv("datafile.csv", header = FALSE)
```

The resulting data frame will have columns named `V1`, `V2`, and so on, and you will probably want to rename them manually:

```{r, eval=FALSE}
# Manually assign the header names
names(data) <- c("Column1", "Column2", "Column3")
```

You can set the delimiter with sep. If it is space-delimited, use `sep = " "`. If it is tab-delimited, use `\t`, as in:

```{r, eval=FALSE}
data <- read.csv("datafile.csv", sep = "\t")
```

By default, strings in the data are treated as factors. Suppose this is your data file, and you read it in using `read.csv()`:

```
"First","Last","Sex","Number"
"Currer","Bell","F",2
"Dr.","Seuss","M",49
"","Student",NA,21
```

The resulting data frame will store `First` and `Last` as *factors*, though it makes more sense in this case to treat them as strings (or *character vectors* in R terminology). To differentiate this, use `stringsAsFactors=FALSE`. If there are any columns that should be treated as factors, you can then convert them individually:

```{r, eval=FALSE}
data <- read.csv("datafile.csv", stringsAsFactors = FALSE)
# Convert to factor
data$Sex <- factor(data$Sex)
str(data)
#> 'data.frame': 3 obs. of 4 variables:
#> $ First : chr "Currer" "Dr." ""
#> $ Last : chr "Bell" "Seuss" "Student"
#> $ Sex : Factor w/ 2 levels "F","M": 1 2 NA
#> $ Number: int 2 49 21
```

**TODO: Fix text output formatting**

Alternatively, you could load the file with strings as factors, and then convert individual columns from factors to characters.

### See Also

`read.csv()` is a convenience wrapper function around `read.table()`. If you need more control over the input, see `?read.table`.


Loading Data from an Excel File
-------------------------------

### Problem

You want to load data from an Excel file.

### Solution

The readxl package has the function `read_excel()` for reading .xls and .xlsx files from Excel. This will read the first sheet of an Excel spreadsheet:

```{r, eval=FALSE}
# Only need to install once
install.packages("readxl")

library(readxl)
data <- read_excel("datafile.xlsx", 1)
```

### Discussion

With `read_excel()`, you can load from other sheets by specifying a number for sheetIndex or a name for sheetName:

```{r, eval=FALSE}
data <- read_excel("datafile.xls", sheet = 2)

data <- read_excel("datafile.xls", sheet = "Revenues")
```

`read_excel()` uses the first row of the spreadsheet for column names. If you don't want to use that row for column names, use `col_names = FALSE`. The columns will instead be named `X1`, `X2`, and so on.

By default, `read_excel()` will infer the type of each column, but if you want to specify the type of each column, you can use the `col_types` argument. You can also drop columns if you specify the type as `"blank"`.

```{r, eval=FALSE}
# Drop the first column, and specify the types of the next three columns
data <- read_excel("datafile.xls", col_types = c("blank", "text", "date", "numeric"))
```


### See Also

See `?read_excel` for more options controlling the reading of these files.

There are other packages for reading Excel files. The gdata package has a function `read.xls()` for reading in .xls files, and the xlsx package has a function `read.xlsx()` for reading in .xlsx files. They require external software to be installed on your computer: `read.xls()` requires Java, and `read.xlsx()` requires Perl.


Loading Data from SPSS/SAS/Stata Files
--------------------------------------

### Problem

You want to load data from a SPSS file, or from other programs like SAS or Stata.

### Solution

The haven package has the function `read_sav()` for reading SPSS files. To load data from an SPSS file:

```{r, eval=FALSE}
# Only need to install the first time
install.packages("haven")

library(foreign)
data <- read_sav("datafile.sav")
```

### Discussion


The haven package also includes functions to read from other formats:

* `read_sas()`: SAS
* `read_dta()`: Stata


An alternative to haven is the foreign package. It also supports SPSS and Stata files, but it is not as up-to-date as the functions from haven. For example, it only supports Stata files up to version 12, while haven supports up to version 14 (the current version as of this writing).

The foreign package does support some other formats, including:

* `read.octave()`: Octave and MATLAB
* `read.systat()`:SYSTAT
* `read.xport()`: SAS XPORT
* `read.dta()`: Stata
* `read.spss()`: SPSS

### See Also

Run `ls("package:foreign")` for a full list of functions in the foreign package.
