Preface {#_preface}
=======

I started using R several years ago to analyze data I had collected for
my research in graduate school. My motivation at first was to escape
from the restrictive environments and canned analyses offered by
statistical programs like SPSS. And even better, because itRSQUOs freely
available, I didnRSQUOt need to convince someone to buy me a copy of the
software—very important for a poor graduate student! As I delved deeper
into R, I discovered that it could also create excellent data graphics.

Each recipe in this book lists a problem and a solution. In most cases,
the solutions I offer arenRSQUOt the only way to do things in R, but
they are, in my opinion, the best way. One of the reasons for RRSQUOs
popularity is that there are many available add-on packages, each of
which provides some functionality for R. There are many packages for
visualizing data in R, but this book primarily uses ggplot2.
(Disclaimer: itRSQUOs now part of my job to do development on ggplot2.
However, I wrote much of this book before I had any idea that I would
start a job related to ggplot2.)

This book isnRSQUOt meant to be a comprehensive manual of all the
different ways of creating data visualizations in R, but hopefully it
will help you figure out how to make the graphics you have in mind. Or,
if youRSQUOre not sure what you want to make, browsing its pages may
give you some ideas about whatRSQUOs possible.

Recipes {#_recipes}
=======

This book is intended for readers who have at least a basic
understanding of R. The recipes in this book will show you how to do
specific tasks. IRSQUOve tried to use examples that are simple, so that
you can understand how they work and transfer the solutions over to your
own problems.

Software and Platform Notes {#_software_and_platform_notes}
===========================

Most of the recipes here use the ggplot2 graphing package. Some of the
recipes require the most recent version of ggplot2, 0.9.3, and this in
turn requires a relatively recent version of R. You can always get the
latest version of R from the [main R project
site](http://www.r-project.org).

> **Note**
>
> If you are not familiar with ggplot2, see [???](#CHAPTER_GGPLOT2) for
> a brief introduction to the package.

Once youRSQUOve installed R, you can install the necessary packages. In
addition to ggplot2, youRSQUOll also want to install the gcookbook
package, which contains data sets for many of the examples in this book.
To install them both, run:

```{r}
install.packages("ggplot2")
install.packages("gcookbook")
```

You may be asked to choose a mirror site for CRAN, the Comprehensive R
Archive Network. Any of the sites should work, but itRSQUOs a good idea
to choose one close to you because it will likely be faster than one far
away. Once youRSQUOve installed the packages, run this in each R session
in which you want to use ggplot2:

```{r}
library(ggplot2)
```

The recipes in this book will assume that youRSQUOve already loaded
ggplot2, so they wonRSQUOt show this line.

If you see an error like this, it means that you forgot to load ggplot2:

    Error: could not find function "ggplot"

The major platforms for R are Mac OS X, Linux, and Windows, and all the
recipes in this book should work on all of these platforms. There are
some platform-specific differences when it comes to creating bitmap
output files, and these differences are covered in
[???](#CHAPTER_OUTPUT).

Conventions Used in This Book {#_conventions_used_in_this_book}
=============================

The following typographical conventions are used in this book:

Italic

:   Indicates new terms, URLs, email addresses, filenames, and file
    extensions.

Constant width

:   Used for program listings, as well as within paragraphs to refer to
    program elements such as variable or function names, databases, data
    types, environment variables, statements, and keywords.

**`Constant width bold`**

:   Shows commands or other text that should be typed literally by the
    user.

*Constant width italic*

:   Shows text that should be replaced with user-supplied values or by
    values determined by context.

> **Tip**
>
> This icon signifies a tip, suggestion, or general note.

> **Caution**
>
> This icon indicates a warning or caution.

Using Code Examples {#_using_code_examples}
===================

This book is here to help you get your job done. In general, you may use
the code in this book in your programs and documentation. You do not
need to contact us for permission unless youRSQUOre reproducing a
significant portion of the code. For example, writing a program that
uses several chunks of code from this book does not require permission.
Selling or distributing a CD-ROM of examples from ORSQUOReilly books
does require permission. Answering a question by citing this book and
quoting example code does not require permission. Incorporating a
significant amount of example code from this book into your
productRSQUOs documentation does require permission.

We appreciate, but do not require, attribution. An attribution usually
includes the title, author, publisher, and ISBN. For example: LDQUO*R
Graphics Cookbook* by Winston Chang (ORSQUOReilly). Copyright 2013
Winston Chang, 978-1-449-31695-2.RDQUO

If you feel your use of code examples falls outside fair use or the
permission given above, feel free to contact us at
permissions@oreilly.com.

> **Note**
>
> Safari Books Online is an on-demand digital library that delivers
> expert content in both book and video form from the world’s leading
> authors in technology and business.

Technology professionals, software developers, web designers, and
business and creative professionals use Safari Books Online as their
primary resource for research, problem solving, learning, and
certification training.

Safari Books Online offers a range of
plans and pricing
for
enterprise
,
government
,
education
, and individuals.
Members have access to thousands of books, training videos, and
prepublication manuscripts in one fully searchable database from
publishers like O’Reilly Media, Prentice Hall Professional,
Addison-Wesley Professional, Microsoft Press, Sams, Que, Peachpit Press,
Focal Press, Cisco Press, John Wiley & Sons, Syngress, Morgan Kaufmann,
IBM Redbooks, Packt, Adobe Press, FT Press, Apress, Manning, New Riders,
McGraw-Hill, Jones & Bartlett, Course Technology, and hundreds
more
. For more information about Safari Books Online, please visit us
online
.
How to Contact Us {#_how_to_contact_us}
=================

Please address comments and questions concerning this book to the
publisher:

O’Reilly Media, Inc.
1005 Gravenstein Highway North
Sebastopol, CA 95472
800-998-9938 (in the United States or Canada)
707-829-0515 (international or local)
707-829-0104 (fax)
We have a web page for this book, where we list errata, examples, and
any additional information. You can access this page at:
<http://oreil.ly/R_Graphics_Cookbook>.

To comment or ask technical questions about this book, send email to
bookquestions@oreilly.com.

For more information about our books, courses, conferences, and news,
see our website at <http://www.oreilly.com>.

Find us on Facebook: <http://facebook.com/oreilly>

Follow us on Twitter: <http://twitter.com/oreillymedia>

Watch us on YouTube: <http://www.youtube.com/oreillymedia>

Acknowledgments {#_acknowledgments}
===============

No book is the product of a single person. There are many people who
helped make this book possible, directly and indirectly. I’d like to
thank the R community for creating R and for fostering a dynamic
ecosystem around it. Thanks to Hadley Wickham for creating the software
that this book revolves around, for pointing O’Reilly in my direction
when they were considering a book about R graphics, and for opening up
many opportunities for me to deepen my knowledge of R.

Thanks to the technical reviewers for this book: Paul Teetor, Hadley
Wickham, Dennis Murphy, and Erik Iverson. Their depth of knowledge and
attention to detail has greatly improved this book. I’d like to thank
the editors at O’Reilly who have shepherded this book along: Mike
Loukides, for guiding me through the early stages, and Courtney Nash,
for pulling me through to the end. I also owe a big thanks to Holly
Bauer and the rest of the production team at O’Reilly, for putting up
with many last-minute edits, and for handling the unusual features of
this book.

Finally, I would like to thank my wife, Sylia, for her support and
understanding—and not just with regard to the book.
